---
title: "Learning R"
author: "Analisa Milkey"
date: "10/20/2022"
output: html_document
---
# set up

```{r}
# load the data into R
# data comes from chapter 1 of Phylogenetic Comparative Methods in R (Revell & Harmon)
trait_data <- read.csv("anole.data.csv") # this dataset has information about trait values for different Anole species
morph_data <- read.csv("ecomorph.csv") # this dataset has information about which species correspond to which ecomorphs (an ecomorph is when different species share the same morphology due to shared habitat)
```

# R Basics

```{r}
# let's view the data that we've read in
View(trait_data)
View(morph_data)

#or just
trait_data
morph_data

# if our data set is large and we just want to view the first few rows, we can use the "Head" function
head(trait_data)
head(morph_data)

# if we want to work with a specific column in a dataset, we can grab it using "$"
trait_data$SVL # this will grab the column that corresponds to the SVL trait

# we can also make this a new object if we want to be able to work with it
SVL_column <- trait_data$SVL
SVL_column

# we can also subset specific rows from a dataset
# in R, [1,] corresponds to the first row of a dataset, while [,1] refers to the first column of a dataset
# let's pull out the first row and give it a name
first_row <- trait_data[1,]
# we can view the object
first_row
# let's pull out the second row too
second_row <- trait_data[2,]
second_row

# let's pull out some columns again in this way
first_column <- trait_data[,1]
first_column
second_column <- trait_data[,2]
second_column

# we can also subset a group of rows or columns like this
first_two_rows <- trait_data[1:2,]
first_two_rows

last_two_columns <- trait_data[,6:7]
last_two_columns

# now let's plot a basic scatterplot with two traits from the trait data spreadsheet, SVL and HL
plot(trait_data$SVL~trait_data$HL)

# let's add axis labels to the scatterplot using xlab and ylab
plot(trait_data$SVL~trait_data$HL, xlab = "HL", ylab="SVL")

# let's also add a title to the scatterplot using main
plot(trait_data$SVL~trait_data$HL, xlab = "HL", ylab="SVL", main = "SVL vs HL")

# we can also add a linear regression line using abline
plot(trait_data$SVL~trait_data$HL, xlab = "HL", ylab="SVL", main = "SVL vs HL")
abline(lm(trait_data$SVL~trait_data$HL))

# we can change the color of the regression line as well
plot(trait_data$SVL~trait_data$HL, xlab = "HL", ylab="SVL", main = "SVL vs HL")
abline(lm(trait_data$SVL~trait_data$HL), col="red")

# we can also change the shape and size of the points on the scatterplot with pch and cex
plot(trait_data$SVL~trait_data$HL, xlab = "HL", ylab="SVL", main = "SVL vs HL", pch=16, cex=0.8)
abline(lm(trait_data$SVL~trait_data$HL), col="red")

plot(trait_data$SVL~trait_data$HL, xlab = "HL", ylab="SVL", main = "SVL vs HL", pch=8, cex=1.0)
abline(lm(trait_data$SVL~trait_data$HL), col="red")

# we can change the color of the points as well
plot(trait_data$SVL~trait_data$HL, xlab = "HL", ylab="SVL", main = "SVL vs HL", pch=16, cex=0.8, col = "blue")
abline(lm(trait_data$SVL~trait_data$HL), col="red")

# now let's try coloring the points by ecomorph type
# we will need to use the other dataset we read in, morph_data
# let's add a new column to morph_data that specifies a color for each ecomorph
# we can add new columns the same way we subset them, using $
# for example, morph_data$test will add a new column named "test" if there isn't one already named that

morph_data$ecomorphcol[morph_data$ecomorph == "TG"] <- "orange" # here, I am telling R to add a new column named "ecomorph" and to add a cell that says "orange" to every row with the "TG" ecomorph
morph_data$ecomorphcol[morph_data$ecomorph == "GB"] <- "green"
morph_data$ecomorphcol[morph_data$ecomorph == "CG"] <- "blue"
morph_data$ecomorphcol[morph_data$ecomorph == "TC"] <- "red"
morph_data$ecomorphcol[morph_data$ecomorph == "Tw"] <- "purple"
morph_data$ecomorphcol[morph_data$ecomorph == "Tr"] <- "pink"

# let's view the new data set
morph_data
#There are lots of color options! For more options, check out a options list at http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf

# now we can color points on the scatterplot by ecomorph type
plot(trait_data$SVL~trait_data$HL, xlab = "HL", ylab="SVL", main = "SVL vs HL", pch=16, cex=0.8, col = morph_data$ecomorphcol)
abline(lm(trait_data$SVL~trait_data$HL), col="red")

# this plot would be more informative with a legend, so let's add one that tells  us what the colors of the points mean
# it's important to make sure the ecomorphs and colors are in the same order as above, otherwise the legend won't match the points
plot(trait_data$SVL~trait_data$HL, xlab = "HL", ylab="SVL", main = "SVL vs HL", pch=16, cex=0.8, col = morph_data$ecomorphcol)
abline(lm(trait_data$SVL~trait_data$HL), col="red")
legend("topleft", c("TG", "GB", "CG", "TC", "Tw", "Tr"), pch=16, pt.cex=1.0, col=c("orange", "green", "blue", "red", "purple", "pink"))

# You can change where the legend in the plot is: e.g., "topright", "bottomleft", etc

# lastly, let's save our plot as a pdf
pdf("SVL_HL.pdf")
plot(trait_data$SVL~trait_data$HL, xlab = "HL", ylab="SVL", main = "SVL vs HL", pch=16, cex=0.8, col = morph_data$ecomorphcol)
abline(lm(trait_data$SVL~trait_data$HL), col="red")
legend("topleft", c("TG", "GB", "CG", "TC", "Tw", "Tr"), pch=16, pt.cex=1.0, col=c("orange", "green", "blue", "red", "purple", "pink"))
dev.off()

# a pdf of the plot should now appear in your working directory


# Questions:
# Try making a scatterplot for two more traits. Can you add an informative x axis, y axis, and title? Does there seem to be a correlation between the traits?
plot(trait_data$LAM~trait_data$FLL, xlab = "FLL", ylab = "LAM", main = "LAM vs FLL")
# Can you color by ecomorph, using different colors?
plot(trait_data$LAM~trait_data$FLL, xlab = "FLL", ylab = "LAM", main = "LAM vs FLL", pch=8, cex=0.8, col = morph_data$ecomorphcol)
# Can you change the shape and size of the points?
plot(trait_data$LAM~trait_data$FLL, xlab = "FLL", ylab = "LAM", main = "LAM vs FLL", pch=32, cex=0.6, col = morph_data$ecomorphcol)
# Can you export this plot as a pdf?
pdf(file = "/Users/ElenaKorte/Desktop/learning-R/LAM_vs_FLL.pdf")
plot(trait_data$LAM~trait_data$FLL, xlab = "FLL", ylab = "LAM", main = "LAM vs FLL")
dev.off()
# What if I wanted to color ecomorphs TG, GB, and CG the same color? How could you go about doing this?
```

# Phylogenetic Independent Contrasts
This idea was introduced in 1985 by Joe Felsenstein. Let's say that I want to figure out if bigger animals are better at retrieving objects than smaller objects. I take a group of dogs and cats and measure how good they are at retrieving objects.  Let's make up some data and analyze it:

## an example
```{r}
# we can create columns of a data set using the c() function
animal <- c("cat", "cat", "cat", "cat", "cat", "cat", "cat", "cat", "dog", "dog", "dog", "dog", "dog", "dog", "dog", "dog")
species <- c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 15, 16)
retrieval_index <- c(0.1, 0.2, 0.15, 0.25, 0.3, 0.27, 0.28, 0.19, 0.9, 0.98, 0.8, 0.85, 0.89, 0.92, 0.93, 0.96)
weight <- c(8, 9, 10, 8, 9, 10, 8.5, 8.9, 60, 75, 70, 71, 66, 75, 72, 73)

# join the variables to create a data frame
dat <- data.frame(animal, retrieval_index, weight, species)


# let's view the data frame
View(dat)

# let's try running a linear regression on retrieval index vs weight
# "lm" stands for "linear model"
lmDat <- lm(dat$retrieval_index~dat$weight)
# view the results
summary(lmDat)

# this summary shows us that there's a significant relationship between retrieval index and weight (p <<< 0.05) (a higher weight is significantly correlated to a higher retrieval index)

# let's view our results as a scatterplot now
plot(dat$retrieval_index~dat$weight, pch=16, xlab = "weight", ylab = "retrieval index")+
  abline(lm(dat$retrieval_index~dat$weight), col="black")

# now let's color the points by type of animal
# create new color column
dat$animal_color[dat$animal == "cat"] <- "green"
dat$animal_color[dat$animal == "dog"] <- "blue"

plot(dat$retrieval_index~dat$weight, col=dat$animal_color, pch=16, xlab = "weight", ylab = "retrieval index")+
  abline(lm(dat$retrieval_index~dat$weight), col="black")+
  legend("topleft", c("cat", "dog"), pch=16, pt.cex=1.0, col=c("green", "blue"))

# it looks like dogs are clustered pretty tightly together, and so are cats
# there are essentially two points in our scatterplot - dogs and cats
# we can't treat these groups as independent
# this is where independent contrasts come in
# we would need the phylogeny of the animals we're studying
# Joe Felsenstein pointed out that while the data between the species are not independent due to shared phylogenetic history, the differences (contrasts) between species are
# to calculate an independent contrast, we would take the difference between two tip values and divide by variance (or use an R package to do it for us!)


# let's correct for phylogeny now
# we will need to read in the dog and cat phylogenetic tree (note that I just made up this tree)
# to read in the tree, we will need to use the package ape
# if you don't have this package installed, you will need to install it using install.packages("ape")
# then load it (you have to load it each time you restart R, but you only have to install it once)
library(ape)
dog_cat_tree<-read.tree("dog_cat_tree.tre")

# to do our analysis, we will also need the package "phytools"
library(phytools)

# let's start by plotting the tree we read in
plotTree(dog_cat_tree)

# let's see if our regression changes when we account for phylogeny
# we need to first pull out the traits as vectors
RI <- setNames(dat[,"retrieval_index"], rownames(dat))
weight <- setNames(dat[,"weight"], rownames(dat))

# compute the independent contrasts for SVL and HL
pic_RI <- pic(RI, dog_cat_tree)
pic_weight <- pic(weight, dog_cat_tree)

# let's now fit a linear regression to these contrasts
pic_lm <- lm(pic_RI~pic_weight+0) # the +0 indicates no intercept (don't worry about this for now)
summary(pic_lm)
# looks like our model is no longer significant (p > 0.05)
# this means that when we take the phylogeny into account, there's no longer a significant correlation between weight and retrieval index

# let's plot our contrasts as well
par(mar=c(5.1,5.1,1.1,1.1)) # this sets the margins so our plot is more readable
plot(pic_RI~pic_weight, xlab = "RI contrasts", ylab = "weight contrasts", pch=16)+
  abline(h=0,lty="dotted")+ # these two lines add grid lines to the plot
  abline(v=0,lty="dotted")+
  abline(pic_lm,lwd=2,col="blue") # add our regression line from our contrasts analysis

# now let's plot both the phylogenetic contrasts and the ordinary regression model
par(mar=c(5.1,5.1,1.1,1.1)) # this sets the margins so our plot is more readable
plot(dat$retrieval_index~dat$weight, xlab = "weight", ylab = "RI", pch=16)+
  abline(h=0,lty="dotted")+ # these two lines add grid lines to the plot
  abline(v=0,lty="dotted")+
  abline(pic_lm,lwd=2,col="blue")+ # add our regression line from our contrasts analysis
  abline(lmDat, lwd=2, col="red")+
  legend("topleft", c("non-phylogenetic regression", "contrasts"), pch=16, pt.cex=1.0, col=c("red", "blue"))
```

## now let's try this with our anole data

```{r}
# let's load in our data again
# we will also need to read in the Anole phylogenetic tree
library(ape)
trait_data <- read.csv("anole.data.csv", row.names=1, header=TRUE)
morph_data <- read.csv("ecomorph.csv", row.names=1, header=TRUE, stringsAsFactors=TRUE)
# the argument row.names=1 tells R the row names are in the first column of the data file
# the argument header=TRUE tells R the first row of the data file contains the column names
anolis_tree<-read.tree("Anolis.tre")

# to do our analysis, we will also need the package "phytools"
library(phytools)

# let's start by plotting the tree we read in
plotTree(anolis_tree)

# this is pretty messy, but we can change the size of the font and the line width to make it a little more readable
plotTree(anolis_tree, fsize=0.5, lwd=1)

# let's start by running an ordinary linear regression on traits SVL and HL without accounting for phylogeny
# we should also log transform our data to account for different body sizes (this attempts to normalize our data & make the statistics more valid)
lmAnole <- lm(log(trait_data$SVL)~log(trait_data$HL))
# view the results
summary(lmAnole)

# this regression is telling us there is a significant correlation between SVL and HL (p < 0.05)

# let's see if this changes when we account for phylogeny
# we need to first pull out SVL and HL as vectors
SVL <- setNames(trait_data[,"SVL"], rownames(trait_data))
HL <- setNames(trait_data[,"HL"], rownames(trait_data))

# compute the independent contrasts for SVL and HL
pic_SVL <- pic(log(SVL), anolis_tree)
pic_HL <- pic(log(HL), anolis_tree)

# let's now fit a linear regression to these contrasts
pic_lmAnole <- lm(pic_SVL~pic_HL+0) # the +0 indicates no intercept (don't worry about this for now)
summary(pic_lmAnole)

# let's plot our contrasts as well
plot(pic_SVL~pic_HL, xlab = "HL contrasts", ylab = "SVL contrasts")+
  abline(h=0,lty="dotted")+ # these two lines add grid lines to the plot
  abline(v=0,lty="dotted")+
  abline(pic_lmAnole,lwd=2,col="blue") # add our regression line from our contrasts analysis

# now let's plot the ordinary regression model
plot(log(trait_data$SVL)~log((trait_data$HL)), xlab = "HL", ylab = "SVL", xlim = c(0, 5.0), ylim = c(0, 2.0))+
  abline(lmAnole, lwd=2, col="red")+ # add regression line from original linear model
  abline(pic_lmAnole, lwd=2, col="blue")+ # add regression line from contrasts analysis
    abline(h=0,lty="dotted")+ # these two lines add grid lines to the plot
  abline(v=0,lty="dotted")+
  legend("bottomright", c("non-phylogenetic regression", "contrasts"), pch=16, pt.cex=1.0, col=c("red", "blue"))

# Questions:
# Try running the same PIC analysis on two more traits from this dataset. Does accounting for phylogeny change the significance of the result?
```


# Principal Components Analysis
PCA is a technique for reducing the dimensionality of datasets, which makes them easier to interpret.

```{r}
# let's read in the data again to start fresh
trait_data <- read.csv("anole.data.csv")
morph_data <- read.csv("ecomorph.csv")

# we are interested in working with the traits, so we can subset those columns
# if we were only interested in a few of the traits, we could subset the data further
trait_data[,2:7]

# run the PCA on columns 2 through 7
pca <- prcomp(trait_data[,2:7])

# extract the PC scores - these scores show us the coefficients for the different PC axes, creating the PCs
scores <- pca$x

# let's check the PC loadings - these loadings show us which traits load most strongly on the different axes of variation
View(pca$rotation)
# The trait FLL seems to load strongly on PC1, as do traits HL and HLL. The trait TL loads strongly negative on PC2. What trait loads the most strongly on PC3?

# now let's plot PC2 vs PC1
plot(PC2 ~ PC1, scores)

# this plot would be more interesting if we could color the points by ecomorph

# let's start by assigning colors to each ecomorph in the ecomorph data file we read in earlier
morph_data$ecomorphcol[morph_data$ecomorph == "TG"] <- "orange"
morph_data$ecomorphcol[morph_data$ecomorph == "GB"] <- "green"
morph_data$ecomorphcol[morph_data$ecomorph == "CG"] <- "blue"
morph_data$ecomorphcol[morph_data$ecomorph == "TC"] <- "red"
morph_data$ecomorphcol[morph_data$ecomorph == "Tw"] <- "purple"
morph_data$ecomorphcol[morph_data$ecomorph == "Tr"] <- "pink"

# let's plot PC2 vs PC1 again, using the colors we just assigned to each ecomorph
plot(PC2 ~ PC1, scores, col=morph_data$ecomorphcol, pch=16, cex=1.0)

# now let's add a legend that tells us what the point colors mean
# make sure ecomorphs and colors are in the same order as above
legend("topright", c("TG", "GB", "CG", "TC", "Tw", "Tr"), pch=16, pt.cex=1.0, col=c("orange", "green", "blue", "red", "purple", "pink"))

# You can change where the legend in the plot is: e.g., "topright", "bottomleft", etc

# let's also add a title

plot(PC2 ~ PC1, scores, col=morph_data$ecomorphcol, pch=16, cex=1.0, main = "Principal Components Analysis")
legend("topright", c("TG", "GB", "CG", "TC", "Tw", "Tr"), pch=16, pt.cex=1.0, col=c("orange", "green", "blue", "red", "purple", "pink"))

# we can also make the axis labels more informative
# Let's make them display the amount of variation contained in that PC axis

summary(PCA) # look at the "proportion of variance" column
# We can see that in this dataset, PC1 explains 90.75% of the variance, and PC2 explains 5.67% of the variance.

# we can add those labels to our plot

plot(PC1 ~ PC2, scores, col=morph_data$ecomorphcol, pch=16, cex=1.0, main = "Principal Components Analysis", xlab = "PC1: 90.75% of variance", ylab = "PC2: 5.67% of variance")
legend("topright", c("TG", "GB", "CG", "TC", "Tw", "Tr"), pch=16, pt.cex=1.0, col=c("orange", "green", "blue", "red", "purple", "pink"))

# let's save our plot as a pdf
pdf("PCA.pdf")
plot(PC2 ~ PC1, scores, col=morph_data$ecomorphcol, pch=16, cex=1.0, main = "Principal Components Analysis", xlab = "PC1: 90.75% of variance", ylab = "PC2: 5.67% of variance")
legend("topright", c("TG", "GB", "CG", "TC", "Tw", "Tr"), pch=16, pt.cex=1.0, col=c("orange", "green", "blue", "red", "purple", "pink"))
dev.off()


# to think about:
# How can we plot PC3 vs PC4? How about PC1 vs PC3?
# What if I only wanted to analyze traits SVL, HL, and HLL in my PCA? How can we do this?
# What traits seem to be contributing most to variance in this dataset?
# Do species of the same ecomorph tend to show similar traits? How can we tell? Did we expect ecomorphs to show similar traits? Why does our plot not show this expected result?
# Can you change the colors we're using in the PCA plot?
```


# Phylogenetic Principal Components Analysis

```{r}
# let's load in our data again
# we will also need to read in the Anole phylogenetic tree
# to read in the tree, we will need to use the package ape
# if you don't have this package installed, you will need to install it using install.packages("ape")
# then load it (you have to load it each time you restart R, but you only have to install it once)
library(ape)
trait_data <- read.csv("anole.data.csv", row.names=1, header=TRUE)
morph_data <- read.csv("ecomorph.csv", row.names=1, header=TRUE, stringsAsFactors=TRUE)
# the argument row.names=1 tells R the row names are in the first column of the data file
# the argument header=TRUE tells R the first row of the data file contains the column names
anolis_tree<-read.tree("Anolis.tre")

# let's look at the tree we just read in
# we can view the tip labels
anolis_tree$tip.label

# we can also view a summary of the tree
anolis_tree

# to view the tree, we can use plotTree
# we will need the package phytools for this (make sure to install it if you don't have it already)
library(phytools)
plotTree(anolis_tree)

# this tree is pretty messy and hard to read, so let's make it more readable
plotTree(anolis_tree, fsize=0.4, lwd=1)

# now let's plot a phylogenetic PCA, where we take into account nonindependence of species when calulating the correlations between different traits
# first, we need to make sure the anole tree and ecomorph dataset have the same number of species
check1 <- name.check(anolis_tree, trait_data)
check1
# the OK tells us that the tree names match the trait data names

check2 <- name.check(anolis_tree, morph_data)
check2
# now we can see that there are some differences between the data and the phylogeny - the ecomorph data has fewer species than the phylogeny

# we need to prune all the taxa that are present in the phylogeny but not in the ecomorph data frame using the function drop.tip
anolis_tree<-drop.tip(anolis_tree, check2$tree_not_data)

# let's double check that this worked
check2 <- name.check(anolis_tree, morph_data)
check2

# now let's drop those names from the trait data by subsetting the taxa we want
trait_data<-trait_data[anolis_tree$tip.label,]

# let's double check that this worked
check3 <- name.check(anolis_tree, trait_data)
check3

# now let's run the phylogenetic PCA
ecomorph_phylo_pca <- phyl.pca(anolis_tree, trait_data)

# view loadings
ecomorph_phylo_pca$V

# extract scores
scores <- ecomorph_phylo_pca$S

# plot the phylogenetic PCA
plot(PC2~PC1, scores)

# add colors to ecomorph data
morph_data$ecomorphcol[morph_data$ecomorph == "TG"] <- "orange"
morph_data$ecomorphcol[morph_data$ecomorph == "GB"] <- "green"
morph_data$ecomorphcol[morph_data$ecomorph == "CG"] <- "blue"
morph_data$ecomorphcol[morph_data$ecomorph == "TC"] <- "red"
morph_data$ecomorphcol[morph_data$ecomorph == "Tw"] <- "purple"
morph_data$ecomorphcol[morph_data$ecomorph == "Tr"] <- "pink"

# color points by ecomorph and add a legend
plot(PC2 ~ PC1, scores, col=morph_data$ecomorphcol, pch=16, cex=1.0, main = "Principal Components Analysis")
legend("topleft", c("TG", "GB", "CG", "TC", "Tw", "Tr"), pch=16, pt.cex=1.0, col=c("orange", "green", "blue", "red", "purple", "pink"))

# this PCA looks more reasonable!
```

